Finiteness in Cubical Type Theory

### 2020-06-30 sections/finiteness-predicates.tex
finished sort proof

\AgdaInductiveConstructor{yes} is a little more interesting: its payload is
\agdaequiv{x}{y}.
How can we prove that the path between \(x\) and \(y\) is a proposition?
It turns out that there is a class of types for which all paths are
propositions: the \emph{sets}.
\begin{agdalisting} \label{isSet}
  \ExecuteMetaData[agda/Snippets/Introduction.tex]{isSet}
\end{agdalisting}
This is the next homotopy level up from the propositions (Equation~\ref{isProp}).
More importantly, there is an important theorem relating to sets which
\emph{also} relates to decidable equality: Hedberg's theorem
\cite{hedbergCoherenceTheoremMartinLof1998}.
\begin{theorem}[Hedberg's Theorem]
  Any type with decidable equality is a set.
\end{theorem} \noindent
And of course we know that \(A\) here has decidable equality: we were just given
two proofs of that fact at the beginning of this proof!
\end{proof}
\subsection{Restrictiveness}
So far our explorations into finiteness predicates have pushed us in the
direction of ``less informative'': however, as mentioned in the introduction, we
can \emph{also} ask how \emph{restrictive} certain predicates are.
Since split enumerability and manifest Bishop finiteness imply each other we
know that there can be no type which satisfies one but not the other.
We also know that manifest Bishop finiteness implies cardinal finiteness, but we
do \emph{not} have a function in the other direction:
\begin{equation}
  \mathcal{C}(A) \rightarrow \AgdaDatatype{\ensuremath{\mathcal{B}}}(A)
\end{equation}
So the question arises naturally: is there a cardinally finite type which is
\emph{not} manifest Bishop finite?

It turns out the answer is no!
\begin{lemma}
  \begin{equation}
    \neg (\Sigma(A : \mathbf{Type}) , \mathcal{C}(A) \times \neg \AgdaDatatype{\ensuremath{\mathcal{B}}}(A))
  \end{equation}
\end{lemma}
\begin{proof}
  We will actually prove a slightly more general statement.
  For any type \(A\), the following holds:
  \begin{equation}
    \neg (\lVert A \rVert \times \neg A)
  \end{equation}
  The solution becomes more clear if we write out the definition of \(\neg\):
  \begin{equation}
    \frac{\lVert A \rVert \;\;\; A \rightarrow \bot }{\bot}
  \end{equation}
  We clearly need to apply a function of type \(A \rightarrow \bot\) to a value
  of type \(\lVert A \rVert\).
  Luckily, this is permissible, as \(\bot\) is a mere proposition.
\end{proof}

\subsection{Going from Cardinal Finiteness to Manifest Bishop Finiteness}
\begin{lemma} \label{manifest-bishop-to-cardinal}
  Any manifest Bishop finite type is cardinal finite.
\end{lemma}
\begin{theorem} \label{cardinal-to-manifest-bishop}
  Any cardinal finite type with a total order is Bishop finite.
\end{theorem}
The proof for this particular theorem is quite involved in the formalisation, so
we only give its sketch here.
First, note that we actually convert to manifest enumerability first: this can
be converted to split enumerability with decidable equality, which is provided
by cardinal finiteness.

Next, we define permutations.
\begin{definition}[List Permutations]
  Two lists are permutations of each other if their membership proofs are all
  equivalent\footnotemark \cite{danielssonBagEquivalenceProofRelevant2012}.
  \begin{equation}
    \mathit{xs} \leftrightsquigarrow \mathit{ys} = \Pi {(x : A)} , x \in \mathit{xs} \simeq x \in \mathit{ys}
  \end{equation}
\end{definition}

\footnotetext{
  The definition in \cite{danielssonBagEquivalenceProofRelevant2012} and our
  formalisation is slightly different: we say permutations are lists with
  \emph{isomorphic} membership proofs.
  The distinction, as it happens, does not affect our work here.
}

Next, we define a sort function which relies on the provided total order.
We further prove the following fact about this sort function:
\begin{equation}
  \Pi(\mathit{xs}, \mathit{ys} : \AgdaDatatype{List}(A)) , \mathit{xs} \leftrightsquigarrow \mathit{ys} \rightarrow \text{sort}(\mathit{xs}) \equiv \text{sort}(\mathit{ys})
\end{equation}

Next, notice that the support lists of any two proofs of manifest Bishop
finiteness must be permutations of each other.
This will allow us to sort the support list of a proof of cardinal finiteness in
a coherently constant (definition~\ref{prop-trunc},
eliminator~\ref{elim-prop-coh}) way, pulling the support list out from the
truncation.
The cover proof emerges naturally from the definition of the permutation.
\section{Manifest Enumerability} \label{manifest-enumerability}
We have now explored quite far in the direction of restricting information:
indeed, cardinal finiteness, a mere proposition, has the minimal amount of
information a predicate can have.
What we haven't seen, however, is a predicate with a different domain to the
others.
Since manifest Bishop finiteness and split enumerability have equal strength
\todo{ref here}, and since we've shown that there is no type which is cardinal
finite but not manifest Bishop finite, we haven't explored any of the
``restrictiveness'' axis from Figure~\ref{finite-classification}.

>>>>

\AgdaInductiveConstructor{yes} is a little more interesting: its payload is
\agdaequiv{x}{y}.
How can we prove that the path between \(x\) and \(y\) is a proposition?
It turns out that there is a class of types for which all paths are
propositions: the \emph{sets}.
\begin{agdalisting} \label{isSet}
  \ExecuteMetaData[agda/Snippets/Introduction.tex]{isSet}
\end{agdalisting}
This is the next homotopy level up from the propositions (Equation~\ref{isProp}).
More importantly, there is an important theorem relating to sets which
\emph{also} relates to decidable equality: Hedberg's theorem
\cite{hedbergCoherenceTheoremMartinLof1998}.
\begin{theorem}[Hedberg's Theorem]
  Any type with decidable equality is a set.
\end{theorem} \noindent
And of course we know that \(A\) here has decidable equality: we were just given
two proofs of that fact at the beginning of this proof!
\end{proof}
\subsection{Restrictiveness}
So far our explorations into finiteness predicates have pushed us in the
direction of ``less informative'': however, as mentioned in the introduction, we
can \emph{also} ask how \emph{restrictive} certain predicates are.
Since split enumerability and manifest Bishop finiteness imply each other we
know that there can be no type which satisfies one but not the other.
We also know that manifest Bishop finiteness implies cardinal finiteness, but we
do \emph{not} have a function in the other direction:
\begin{equation}
  \agdacal{C}\;A\rightarrow\agdacal{B}\;A
\end{equation}
So the question arises naturally: is there a cardinally finite type which is
\emph{not} manifest Bishop finite?

It turns out the answer is no!
We can prove the following statement:
\begin{agdalisting}
  \ExecuteMetaData[agda/Cardinality/Finite/Cardinal.tex]{no-gap-card-bishop}
\end{agdalisting}
We use the fact that \agdabot\; is itself a proposition to eliminate from under
the truncation.
\begin{agdalisting}
  \ExecuteMetaData[agda/Cardinality/Finite/Cardinal.tex]{no-gap-card-bishop-proof}
\end{agdalisting}

This tells us that manifest bishop finiteness, cardinal finiteness, and split
enumerability all refer to the same class of types.
\subsection{Going from Cardinal Finiteness to Manifest Bishop Finiteness}
We know of course that we can convert any proof of manifest Bishop finiteness to
a proof of Cardinal finiteness: it's just the truncation function
\AgdaInductiveConstructor{\(\lvert \_ \rvert\)}.
It's the other direction which presents a difficulty:
\begin{theorem} \label{cardinal-to-manifest-bishop}
  Any cardinal finite type with a total order is Bishop finite.
\end{theorem}
The proof for this particular theorem is quite involved in the formalisation, so
we only give its sketch here.

Our strategy will be to \emph{sort} the support list of the proof for Bishop
finiteness, and then prove that the sorting function is coherently constant,
thereby satisfying the eliminator in Equation~\ref{elim-prop-coh}.
We need to show, in other words, that sorting two support lists from proofs of
manifest Bishop finiteness on the same type with the same order always returns
the same result.
The sort we use is a standard insertion sort:
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/List/Sort.tex]{insert}
  \end{agdalisting} \columnbreak \null \vfill \null \vfill
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/List/Sort.tex]{insertion-sort}
  \end{agdalisting} \vfill \null
\end{multicols} \vspace{-\baselineskip} \noindent
And we prove that \AgdaFunction{sort} produces a list which is sorted, and a
permutation of its input.
\begin{multicols}{2}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/List/Sort.tex]{sort-sorts}
  \end{agdalisting} \columnbreak
  \begin{agdalisting}
    \ExecuteMetaData[agda/Data/List/Sort.tex]{sort-perm}
  \end{agdalisting}
\end{multicols} \vspace{-\baselineskip}
We've introduced permutations here: we take their definition from
\cite{danielssonBagEquivalenceProofRelevant2012}.
Two lists are permutations of each other if their membership proofs are all
equivalent.
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/List/Relation/Binary/Permutation.tex]{perm-def}
\end{agdalisting}
This definition fits particularly well for two reasons: first, it is defined on
containers generically, which fits well with our finiteness predicates.
Secondly, it is extremely straightforward to show that the support lists of any
two proofs of manifest Bishop finiteness must be permutations of each other:
\begin{agdalisting}
  \ExecuteMetaData[agda/Cardinality/Finite/Cardinal.tex]{perm-bish}
\end{agdalisting}

Almost all of the pieces are in place now: we know that the support lists of all
proofs of \(\agdacal{B}\;A\) are permutations of each other, and we know that
\AgdaFunction{sort} returns a sorted permutation of its input.
The final piece of the puzzle is the following:
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/List/Sort.tex]{sorted-perm-eq}
\end{agdalisting}
If two sorted lists are both permutations of each other they must be equal.
Connecting up all the pieces we get the following:
\begin{agdalisting}
  \ExecuteMetaData[agda/Data/List/Sort.tex]{perm-invar}
\end{agdalisting}
Because we know that all support lists of \(\agdacal{B}\;A\) are permutations of
each other this is enough to prove that \AgdaFunction{sort} is coherently
constant, and therefore can eliminate from within a truncation.
The second component of the output pair (the cover proof) follows quite
naturally from the definition of permutations.
\section{Manifest Enumerability} \label{manifest-enumerability}
We have now explored quite far in the direction of restricting information:
indeed, cardinal finiteness, a mere proposition, has the minimal amount of
information a predicate can have.
What we haven't seen, however, is a predicate with a different domain to the
others.
Since manifest Bishop finiteness and split enumerability have equal strength
\todo{ref here}, and since we've shown that there is no type which is cardinal
finite but not manifest Bishop finite, we haven't explored any of the
``restrictiveness'' axis from Figure~\ref{finite-classification}.
